---
title: " Reproducible Research: Peer Assessment 1 "
output: html_document
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE,message = FALSE)
```
## Loading and preprocessing the data
We first save the activity data into our working directory, load the data into R-studio and format the date column as a date. 
```{r }
library(dplyr)
activity_data=read.csv("activity.csv")
activity_data=as_tibble(activity_data)
activity_data$date=as.Date(activity_data$date, "%Y-%m-%d")
```
We now display the histogram for the total numbers of daily steps.

```{r }
library(dplyr)
 daily_data=activity_data%>%group_by(date)%>%summarize(daily_steps=sum(steps,na.rm=TRUE))
 library(ggplot2)
 g=ggplot(daily_data,aes(x=date, y=daily_steps))
 g+geom_col() + labs(title="Total steps by day", y="Count")
```
  
##  What are the mean and median of total number of steps taken per day?
```{r}
mean_daily_steps=mean(daily_data$daily_steps)
median_daily_steps=median(daily_data$daily_steps)
cat("The average number of daily steps is",mean_daily_steps, "\n")
cat("The median number of daily steps is", median_daily_steps ,"\n")
```
      
## What is the average daily activity pattern?    
Below is the code that pull the data of average daily steps by interval and displays it as a time-series.
```{r}
library(dplyr)
 interval_avg_data=activity_data%>%group_by(interval)%>%summarize(avg_steps=mean(steps,na.rm=TRUE))
 library(ggplot2)
 g=ggplot(interval_avg_data,aes(x=interval, y=avg_steps))
 g+geom_line() + labs(title="Average daily steps", y="Average Steps")

```  
Let us now find the interval with the  the highest daily average steps:
```{r}
target=interval_avg_data$interval[interval_avg_data$avg_steps==max(interval_avg_data$avg_steps)]
cat("The interval with the highest average daily steps is the one starting at",target,"\n")
```  
## Imputing the Missing Data
We aim to replace the missing data with the daily average corresponding to that interval.
The strategy is to merge the original data with the summary data on interval averaged and to override the NAs with the average number of steps.
```{r}
new_activity_data=merge(activity_data,interval_avg_data, by.x="interval", by.y="interval")
new_steps=NULL
for (k in 1:length(new_activity_data$date)){
if (is.na(new_activity_data$steps[k])){new_steps[k]=new_activity_data$avg_steps[k]}
  else {new_steps[k]=new_activity_data$steps[k]}
}
activity_data_new=cbind(new_activity_data[,c(1,3)],new_steps)
```
  
## Histogram of the total dailty steps after imputing


Plotting the updated data as a time-series.
```{r}
 
library(dplyr)
 daily_data_new=activity_data_new%>%group_by(date)%>%summarize(new_daily_steps=sum(new_steps,na.rm=TRUE))
 library(ggplot2)
 g=ggplot(daily_data_new,aes(x=date, y=new_daily_steps))
 g+geom_col() + labs(title="Total steps by day", y="Count")
```
  
##  For comparison, What are the mean and median of total number of steps taken per day?
```{r}
mean_daily_steps_new=mean(daily_data_new$new_daily_steps)
median_daily_steps_new=median(daily_data_new$new_daily_steps)
cat("The average number of daily steps is",mean_daily_steps_new, "\n")
cat("The median number of daily steps is", median_daily_steps_new ,"\n")
```
  
## Preparing the data for the panel plot of weekdays vs weekends
```{r}
weekday=c("Monday","Tuesday","Wednesday","Thursday", "Friday")
status=NULL
for(k in 1:length(activity_data_new$date)){
    if (weekdays(activity_data_new$date[k]) %in% weekday){  status[k]="Weekday"}
    else  {status[k]="Weekend"}
}
Status=as.factor(status)
activity_data_final=cbind(activity_data_new, Status)%>%group_by(Status,interval)%>% summarize(avg_steps=mean(new_steps))
```  

## The panel plot of Weekday vs Weekend

```{r}
g2=ggplot(activity_data_final,aes(x=interval, y=avg_steps))
 g2+geom_line() + facet_wrap(~Status,ncol=1)+labs(title="Average daily steps Weekdays vs weekends", y="Average Steps")
```




 